MSPMS-Shiny

### 2024-04-16 README.Rmd
major changes

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

@@

1. Takes two input files from PEAKS and combines them
3. Normalizes values and then does a reverse log2 transformation
4. Looks for outliers across replicates. Removes them. 
5. Imputes data for missing values (likely to be very low intesity).
6. Figures out the locations of the detected clevages within the library of peptide sequences used. Is it cleaved at the N or C terminus, or both?
7. Calculates the fold change and the p/q value across experimental conditions using T-tests.

@@

The files coming from peaks should have..... **Describe how to do this here**.

Importantly, this. data contains... what????

@@

### Loading the files ###
lfq_filename = "legacy/protein-peptides-lfq.csv"
#file "protein-peptides.csv" exported from PEAKS identification
id_filename = "legacy/protein-peptides-id.csv"

# Prepare the data for normalyzer analysis
prepared_data = prepare_for_normalyzer(lfq_filename,id_filename)
```

@@

Before we normalyze data, we need to know what samples are in what groups. We can do that by defining a design matrix. 
```{r}
design_matrix = readr::read_csv("annotation.csv")

@@

Now we can normalyze the data.
```{r}
normalyzed_data = normalyze(prepared_data,design_matrix)
```

@@

```{r}
design_matrix = readr::read_csv("annotation.csv")
outliers = handle_outliers(normalyzed_data,design_matrix)
```

@@

```{r}
imputed = impute(outliers)
```

@@

Next we need to join everything with the sequences of the peptide library
```{r}
joined_with_library = join_with_library(imputed)
```

### Calcuating clevages
Next, we need to determine where the peptide sequences are cleaved.

@@

```{r}
final_data = add_cleavages(joined_with_library)

head(final_data)
```

@@

Right now this is left to the user to do on their own. Will update after talkng to Bri.

>>>>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
```

@@

1. Takes two input files from PEAKS and combines them
3. Normalizes values and then does a reverse log2 transformation
4. Looks for outliers across replicates. Removes them. 
5. Imputes data for missing values (likely to be very low intesity).
6. Figures out the locations of the detected clevages within the library of peptide sequences used. Is it cleaved at the N or C terminus, or both?

This package also contains tools for doing some basic statistics and data visualizations.

7. Calculates the fold change and the p/q value across experimental conditions using T-tests.   
8. Conducts ANOVA tests to determine if there are any peptides that are significantly different across all time points.  
9. Conducts a hierarchical clustering analysis to visualize overall patterns in the data.

@@

All analysis performed by this package is downstream of the data generated by the PEAKS software. The first step is to combine the two files generated by PEAKS.

The files coming from peaks should be generated according to the instructions found [here](www/PeaksDataAnalysis_howto.pdf).

@@

### Loading the files ###
lfq_filename = "test/protein-peptides-lfq.csv"
#file "protein-peptides.csv" exported from PEAKS identification
id_filename = "test/protein-peptides-id.csv"

# Prepare the data for normalyzer analysis
prepared_data = prepare_for_normalyzer(lfq_filename,id_filename)

# saving to data folder in package for tests. Ignore when using on your own data
usethis::use_data(prepared_data,overwrite = TRUE)
```

@@

Before we normalyze data, we need to know what samples are in what groups. We can do that by defining a design matrix. 
```{r}
design_matrix = readr::read_csv("test/design_matrix.csv")

@@

Now we can normalyze the data.
```{r}
normalyzed_data = normalyze(prepared_data,design_matrix)

# saving to data folder in package for tests. Ignore when using on your own data
usethis::use_data(normalyzed_data,overwrite = TRUE)
```

@@

```{r}
design_matrix = readr::read_csv("test/design_matrix.csv")
outliers = handle_outliers(normalyzed_data,design_matrix)

# saving to data folder in package for tests. Ignore when using on your own data
usethis::use_data(outliers,overwrite = TRUE)
```

@@

```{r}
imputed = impute(outliers)

# saving to data folder in package for tests. Ignore when using on your own data
usethis::use_data(imputed,overwrite = TRUE)
```

@@

Next we need to join everything with the sequences of the peptide library
```{r}
joined_with_library = join_with_library(imputed)

# saving to data folder in package for tests. Ignore when using on your own data
usethis::use_data(joined_with_library,overwrite = TRUE)
```

### Calcuating clevages.

Next, we need to determine where the peptide sequences are cleaved.

@@

```{r}
cleavage_added_data = add_cleavages(joined_with_library)

head(cleavage_added_data)

# saving to data folder in package for tests. Ignore when using on your own data
usethis::use_data(cleavage_added_data,overwrite = TRUE)
```

### polishing

Someimes there is a need to polish the data a bit for downstream analysis. This function does that by removing combining the cterm and nterm cleavage information into one column while removing any rows that don't have any cleavage information or have cleavage information on the cterm and nterm.

```{r}
polished_data = polish(cleavage_added_data)
```

@@

mspms provides a number of convenience functions to conduct statistics on the data.

First, we need to prepare the data for stats. This involves reshaping the data into the long format so it is easy to conduct statistics on and then appending the data in the design matrix that allows us to determine what sample contains what conditions/time and conduct the appropriate statistics.

```{r}
prepared_for_stats = prepare_for_stats(cleavage_added_data,design_matrix)

usethis::use_data(prepared_for_stats,overwrite = TRUE)
```

Noe we can conduct the statistics.

#### T tests

Ttests are performed within each condition and compared to time 0.

For example, for an experiment where there are two conditions, DMSO and MZB as well as 3 time points, 0, 1, and 2, the t tests would be as follows:

* DMSO.TO vs DMSO.T0. 
* DMSO.T0 vs DMSO.T1. 
* DMSO.T0 vs DMSO.T2.

* MZB.T0 vs MZB.T0. 
* MZB.T0 vs MZB.T1. 
* MZB.T0 vs MZB.T2.

```{r}
# Perform T test
t_test_stats = mspms::mspms_t_tests(prepared_for_stats)
```

#### log2FC

We can also calculate the log 2 fc. The comparisons here are the same as for the T tests
```{r}
# calclulate log2fc
log2fc = mspms::mspms_log2fc(prepared_for_stats)

head(log2fc)
```

#### log2fc_t_tests. 
Sometimes (such as when you want to make volcano plots) it is useful to look at the log2fc and the t test statistics at the same time. This function does that.

```{r}
log2fc_t_test = mspms::log2fc_t_test(prepared_for_stats)

head(log2fc_t_test)
```

Now that we have this data we can visualize these results easily using the ggplot2 package.

```{r}
library(ggplot2)

p1 = log2fc_t_test %>% 
  ggplot(aes(x = log2fc,y = -log10(p.adj)))+
  geom_point(size =0.5)+
  geom_hline(yintercept = -log10(0.05),linetype = "dashed",color = "red")+
  geom_vline(xintercept = 3, linetype = "dashed",color = "red")+
  geom_vline(xintercept = -3, linetype = "dashed",color = "red")+
  theme_minimal()+
  labs(x = "Log2 Fold Change",y = "-log10(p value)")+
  facet_wrap(~comparison,scales = "free")

p1
```

#### ANOVA

We also might want to perfom an anova. Here, we have it set up to show the effect of time within each condition.

For example, for an experiment where there are two conditions, DMSO and MZB as well as 3 time points, 0, 1, and 2, the anova would statistics for the effect of time for each peptide within DMSO or MZB.

```{r}
# Doing ANOVA
anova_stats = mspms::mspms_anova(prepared_for_stats)

head(anova_stats)
```

### Common Data Visualizations

We also provide some functions that make common data visualizations easier.

#### PCA

We can generate a PCA plot to visualize the data. Here, the colors show the different time points while the shape shows the different conditions.

```{r}

mspms::plot_pca(prepared_for_stats)
```

#### Hierchical clustering

We can also generate an interactive heatmap with the hierchical clustering results.

The format of this readme does not allow for interactivity, so a static picture of the output is shown instead.

```{r,eval = FALSE}

mspms::plot_heatmap(prepared_for_stats)

```

![](www/plot_heatmap_output.png)

### 2024-04-16 mspms_shiny/about.Rmd
major changes

On the side bar, you will see two places to upload files.

1. The first file should be the LFQ file from PEAKS. This file should be a csv file.
2. The second file should be the ID file from PEAKS. This file should also be a csv file.

>>>>

On the side bar, you will see three places to upload files.

1. The first file should be the LFQ file as exported from PEAKS. This file should be a csv file.

This file should look like the following:

```{r}
DT::datatable(head(mspms::peaks_lfq,2))
```

2. The second file should be the ID file as exported from PEAKS. This file should also be a csv file.

This file should look like the following:

```{r}
DT::datatable(head(mspms::peaks_id,2))
```

3. The third file should be the design matrix file. This should be a csv file.

The design matrix file should be a csv file with the following columns: "sample" and "group". The sample column should contain the names of the samples as found in the PEAKS files. The group column should contain the group that the sample belongs to. This can be any thing, as long as it is the same per group.

Here is an example of what the design matrix file should look like for the test files.

```{r}
DT::datatable(mspms::design_matrix)
```
